---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# paulmisc

<!-- badges: start -->
[![R-CMD-check](https://github.com/prcleary/paulmisc/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/prcleary/paulmisc/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

> Collection of miscellaneous R functions of interest only to Paul

This package provides tools for epidemiological visualization and data simulation, with a focus on outbreak analysis and epidemic curves.

## Installation 

Install the development version from GitHub:

```{r eval=FALSE}
# Install remotes if you don't have it
# install.packages("remotes")

remotes::install_github('prcleary/paulmisc')
```

## Features

- **`geom_epicurve()`** - A ggplot2 geom for creating classical epidemic curves where each case is represented as a small square
- **`simulate_outbreak()`** - Generate realistic outbreak data for testing and examples

## Usage

```{r setup, message=FALSE, warning=FALSE}
# Load package from source for rendering
devtools::load_all()
library(ggplot2)
```

### Basic Epidemic Curve

Create a simple epidemic curve from simulated outbreak data:

```{r basic-epicurve, fig.height=4, fig.width=7}
# Simulate a point-source outbreak
cases <- simulate_outbreak(n = 50, seed = 42)

# Create basic epicurve
ggplot(cases, aes(x = onset_date)) +
  geom_epicurve() +
  labs(
    title = "Outbreak Epicurve",
    x = "Date of Onset",
    y = "Number of Cases"
  ) +
  theme_minimal()
```

### Colored by Category

Visualize cases by demographic or clinical characteristics:

```{r colored-epicurve, fig.height=4, fig.width=7}
# Color by age group
ggplot(cases, aes(x = onset_date, fill = age_group)) +
  geom_epicurve(colour = "grey20") +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Cases by Age Group",
    x = "Date of Onset",
    y = "Number of Cases",
    fill = "Age Group"
  ) +
  theme_bw()
```

### Faceted Analysis

Compare outbreaks across different settings or groups:

```{r faceted-epicurve, fig.height=6, fig.width=7}
# Facet by setting and color by outcome
ggplot(cases, aes(x = onset_date, fill = outcome)) +
  geom_epicurve(height = 0.85) +
  facet_wrap(~ setting, ncol = 1, scales = "free_y") +
  scale_fill_manual(
    values = c("Recovered" = "steelblue", "Hospitalised" = "tomato")
  ) +
  labs(
    title = "Outbreak Comparison by Setting",
    x = "Date of Onset",
    y = "Number of Cases",
    fill = "Outcome"
  ) +
  theme_minimal() +
  theme(panel.spacing = unit(1, "lines"))
```

### Custom Incubation Periods

Simulate outbreaks with different epidemiological characteristics:

```{r incubation-comparison, fig.height=6, fig.width=7}
# Short incubation period (e.g., food poisoning)
short_incubation <- simulate_outbreak(
  n = 100,
  exposure = as.Date("2024-08-15"),
  meanlog = 0.5,   # ~1.6 day median
  sdlog = 0.3,
  seed = 123
)

# Long incubation period (e.g., viral hepatitis)
long_incubation <- simulate_outbreak(
  n = 100,
  exposure = as.Date("2024-08-15"),
  meanlog = 3,     # ~20 day median
  sdlog = 0.5,
  seed = 123
)

# Compare side by side
library(patchwork)

p1 <- ggplot(short_incubation, aes(x = onset_date)) +
  geom_epicurve(fill = "coral") +
  labs(title = "Short Incubation", x = NULL, y = "Cases") +
  theme_minimal()

p2 <- ggplot(long_incubation, aes(x = onset_date)) +
  geom_epicurve(fill = "skyblue") +
  labs(title = "Long Incubation", x = "Date of Onset", y = "Cases") +
  theme_minimal()

p1 / p2
```

### Advanced Customization

Fine-tune the appearance of individual case squares:

```{r advanced-styling, fig.height=4, fig.width=7}
# Adjust spacing and size
ggplot(cases, aes(x = onset_date, fill = sex)) +
  geom_epicurve(
    width = 0.8,   # Horizontal spacing (0-1)
    height = 0.95, # Vertical spacing (0-1, higher = less gap)
    colour = "white",
    linewidth = 0.2
  ) +
  scale_fill_viridis_d(option = "plasma", begin = 0.2, end = 0.8) +
  labs(
    title = "Cases by Sex with Custom Styling",
    x = "Date of Onset",
    y = "Number of Cases"
  ) +
  theme_dark()
```

## Development

### Setup

Clone the repository and install development dependencies:

```{r eval=FALSE}
# Install development packages
install.packages(c("devtools", "testthat", "roxygen2", "pkgdown"))

# Load the package for development
devtools::load_all()
```

### Testing

Run the test suite to ensure everything works correctly:

```{r eval=FALSE}
# Run all tests
devtools::test()

# Run tests with coverage report
covr::package_coverage()
```

### Documentation

Update documentation after modifying roxygen comments:

```{r eval=FALSE}
# Generate documentation from roxygen comments
devtools::document()

# Preview documentation for a function
?geom_epicurve
```

### Package Checks

Run R CMD check to ensure the package meets CRAN standards:

```{r eval=FALSE}
# Run comprehensive package checks
devtools::check()

# Check for common issues
goodpractice::gp()
```

### Building the Package

Build and install the package locally:

```{r eval=FALSE}
# Build source package
devtools::build()

# Install from local source
devtools::install()

# Build and reload in current session
devtools::load_all()
```

### Website

This package uses pkgdown for documentation website generation:

```{r eval=FALSE}
# Build the website
pkgdown::build_site()

# Preview locally
pkgdown::preview_site()
```

## Contributing

This is a personal utility package, but suggestions and contributions are welcome through GitHub issues and pull requests.

## License

MIT + file LICENSE
