---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->
<!-- Test: workflow should regenerate this -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# paulmisc

<!-- badges: start -->
[![R-CMD-check](https://github.com/prcleary/paulmisc/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/prcleary/paulmisc/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/prcleary/paulmisc/graph/badge.svg)](https://app.codecov.io/gh/prcleary/paulmisc)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![License: MIT](https://img.shields.io/badge/License-MIT-yellow.svg)](https://opensource.org/licenses/MIT)
<!-- badges: end -->

> Collection of miscellaneous R functions of interest only to Paul

This package provides tools for epidemiological visualisation, data
simulation, and an interactive Shiny app for building SQL queries for
Amazon Redshift.

**Documentation:** https://prcleary.github.io/paulmisc/

## Installation 

### System Dependencies (Linux)

On Linux, you may need to install system libraries required by
ggplot2 and other dependencies:

**Debian/Ubuntu:**
```bash
sudo apt-get install -y \
  libcurl4-openssl-dev \
  libssl-dev \
  libxml2-dev \
  libfontconfig1-dev \
  libharfbuzz-dev \
  libfribidi-dev \
  libfreetype6-dev \
  libpng-dev \
  libtiff5-dev \
  libjpeg-dev
```

**RHEL/Fedora/Rocky Linux:**
```bash
sudo dnf install -y \
  libcurl-devel \
  openssl-devel \
  libxml2-devel \
  fontconfig-devel \
  harfbuzz-devel \
  fribidi-devel \
  freetype-devel \
  libpng-devel \
  libtiff-devel \
  libjpeg-turbo-devel
```

### R Package Installation

Install the development version from GitHub:

```{r eval=FALSE}
# Install remotes if you don't have it
# install.packages("remotes")

remotes::install_github('prcleary/paulmisc')
```

## Features

### Epidemiological Tools

- **`geom_epicurve()`** - A ggplot2 geom for creating classical
  epidemic curves where each case is represented as a small square
- **`simulate_outbreak()`** - Generate realistic outbreak data for
  testing and examples

### Shiny Applications

- **`run_redshift_query_builder()`** - Interactive Shiny app for
  building Amazon Redshift SQL queries without writing code. Features
  include:
  - Form-based query construction
  - Support for WHERE conditions, date filters, aggregates, GROUP BY,
    ORDER BY
  - Real-time validation and error checking
  - One-click copy to clipboard
  - Dark-themed, modern UI

## Usage

### Epidemiological Visualisation

```{r setup, message=FALSE, warning=FALSE}
library(paulmisc)
library(ggplot2)
```

### Basic Epidemic Curve

Create a simple epidemic curve from simulated outbreak data:

```{r basic-epicurve, fig.height=4, fig.width=7}
# Simulate a point-source outbreak
cases <- simulate_outbreak(n = 50, seed = 42)

# Create basic epicurve
ggplot(cases, aes(x = onset_date)) +
  geom_epicurve() +
  labs(
    title = "Outbreak Epicurve",
    x = "Date of Onset",
    y = "Number of Cases"
  ) +
  theme_minimal()
```

### Coloured by Category

Visualise cases by demographic or clinical characteristics:

```{r coloured-epicurve, fig.height=4, fig.width=7}
# Colour by age group
ggplot(cases, aes(x = onset_date, fill = age_group)) +
  geom_epicurve(colour = "grey20") +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Cases by Age Group",
    x = "Date of Onset",
    y = "Number of Cases",
    fill = "Age Group"
  ) +
  theme_bw()
```

### Faceted Analysis

Compare outbreaks across different settings or groups:

```{r faceted-epicurve, fig.height=6, fig.width=7}
# Facet by setting and colour by outcome
ggplot(cases, aes(x = onset_date, fill = outcome)) +
  geom_epicurve(height = 0.85) +
  facet_wrap(~ setting, ncol = 1, scales = "free_y") +
  scale_fill_manual(
    values = c("Recovered" = "steelblue", "Hospitalised" = "tomato")
  ) +
  labs(
    title = "Outbreak Comparison by Setting",
    x = "Date of Onset",
    y = "Number of Cases",
    fill = "Outcome"
  ) +
  theme_minimal()
```

### Custom Incubation Periods

Simulate outbreaks with different epidemiological characteristics by
adjusting the incubation period parameters. The `meanlog` parameter
controls the median incubation period (median = `exp(meanlog)` days),
while `sdlog` controls the spread around that median:

```{r incubation-comparison, fig.height=6, fig.width=7}
# Short incubation period (e.g., Salmonella, norovirus)
# Median incubation: exp(0.5) ≈ 1.6 days
short_incubation <- simulate_outbreak(
  n = 100,
  exposure = as.Date("2024-08-15"),
  meanlog = 0.5,
  sdlog = 0.3,
  seed = 123
)

# Long incubation period (e.g., Hepatitis A)
# Median incubation: exp(3) ≈ 20 days
long_incubation <- simulate_outbreak(
  n = 100,
  exposure = as.Date("2024-08-15"),
  meanlog = 3,
  sdlog = 0.5,
  seed = 123
)

# Compare side by side
library(patchwork)

p1 <- ggplot(short_incubation, aes(x = onset_date)) +
  geom_epicurve(fill = "coral") +
  labs(title = "Short Incubation", x = NULL, y = "Cases") +
  theme_minimal()

p2 <- ggplot(long_incubation, aes(x = onset_date)) +
  geom_epicurve(fill = "skyblue") +
  labs(title = "Long Incubation", x = "Date of Onset", y = "Cases") +
  theme_minimal()

p1 / p2
```

### Advanced Customisation

Fine-tune the appearance of individual case squares:

```{r advanced-styling, fig.height=4, fig.width=7}
# Adjust spacing and size
ggplot(cases, aes(x = onset_date, fill = sex)) +
  geom_epicurve(
    width = 0.8,   # Horizontal spacing (0-1)
    height = 0.95, # Vertical spacing (0-1, higher = less gap)
    colour = "white",
    linewidth = 0.2
  ) +
  scale_fill_manual(values = c("Male" = "#0072B2", "Female" = "#D55E00")) +
  labs(
    title = "Cases by Sex with Custom Styling",
    x = "Date of Onset",
    y = "Number of Cases"
  ) +
  theme_minimal()
```

### Interactive Plotly Visualisation

Create interactive epidemic curves with custom tooltips using `plotly`. Note that custom ggplot2 geoms are not compatible with `ggplotly()` conversion, so we use plotly's native functions:

```{r eval=FALSE}
library(plotly)

# Compute stacking positions
cases <- cases[order(cases$onset_date), ]
cases$y <- ave(seq_len(nrow(cases)), cases$onset_date, FUN = seq_along)

# Add custom tooltip text
cases$tooltip <- paste0(
  "Case ID: ", cases$case_id, "<br>",
  "Date: ", cases$onset_date, "<br>",
  "Age: ", cases$age_group, "<br>",
  "Sex: ", cases$sex, "<br>",
  "Setting: ", cases$setting
)

# Create interactive plotly plot
plot_ly(cases) %>%
  add_trace(
    type = "scatter",
    mode = "markers",
    x = ~onset_date,
    y = ~y,
    color = ~age_group,
    marker = list(
      symbol = "square",
      size = 20,
      line = list(color = "white", width = 1)
    ),
    text = ~tooltip,
    hovertemplate = "%{text}<extra></extra>"
  ) %>%
  layout(
    title = "Interactive Epidemic Curve",
    xaxis = list(title = "Date of Onset"),
    yaxis = list(title = "Number of Cases")
  )
```

Users can hover over individual case squares to see detailed information.

**See the [Interactive Epidemic Curves with Plotly](https://prcleary.github.io/paulmisc/articles/interactive-epicurves.html) article for live interactive examples you can try in your browser!**


### Redshift SQL Query Builder

Launch the interactive Shiny app for building SQL queries:

```{r eval=FALSE}
# Launch the Redshift SQL Query Builder Shiny app
run_redshift_query_builder()
```

The app provides a user-friendly interface for:

- **Table Selection**: Specify schema, table name, and optional alias
- **Column Selection**: Choose all columns, specific columns with
  DISTINCT, or aggregate functions (COUNT, SUM, AVG, MIN, MAX, COUNT
  DISTINCT)
- **WHERE Conditions**: Add up to 3 conditions with AND/OR logic using
  various operators (=, !=, >, <, >=, <=, LIKE, ILIKE, IN, NOT IN, IS
  NULL, IS NOT NULL, BETWEEN)
- **Date Filters**: Filter by date ranges, last N days, current
  month/year, or specific dates using Redshift-specific functions like
  DATEADD and TRUNC
- **Sorting & Grouping**: GROUP BY, HAVING, ORDER BY with LIMIT and OFFSET
- **Validation**: Real-time error checking with helpful validation messages
- **Copy to Clipboard**: One-click copy of the generated SQL query

The app features a modern dark theme and includes helpful Redshift
SQL tips for common functions and patterns.

## Development

### Setup

Clone the repository and install development dependencies:

```{r eval=FALSE}
# Install development packages
install.packages(c("devtools", "testthat", "roxygen2", "pkgdown"))

# Load the package
library(paulmisc)
```

### Testing

Run the test suite to ensure everything works correctly:

```{r eval=FALSE}
# Run all tests
devtools::test()

# Run tests with coverage report
covr::package_coverage()
```

### Documentation

Update documentation after modifying roxygen comments:

```{r eval=FALSE}
# Generate documentation from roxygen comments
devtools::document()

# Preview documentation for a function
?geom_epicurve
```

### Package Checks

Run R CMD check to ensure the package meets CRAN standards:

```{r eval=FALSE}
# Run comprehensive package checks
devtools::check()

# Check for common issues
goodpractice::gp()
```

### Building the Package

Build and install the package locally:

```{r eval=FALSE}
# Build source package
devtools::build()

# Install from local source
devtools::install()

# Or load with library
library(paulmisc)
```

### Website

This package uses pkgdown for documentation website generation:

```{r eval=FALSE}
# Build the website
pkgdown::build_site()

# Preview locally
pkgdown::preview_site()
```

## Contributing

This is a personal utility package, but suggestions and contributions
are welcome through GitHub issues and pull requests.

## License

MIT + file LICENSE
