---
title: "Interactive Epidemic Curves with Plotly"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Interactive Epidemic Curves with Plotly}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Overview

The `geom_epicurve()` function works seamlessly with [plotly](https://plotly.com/r/) to create interactive epidemic curves with custom tooltips. Simply create your plot with ggplot2 and convert it using `ggplotly()`.

## Basic Interactive Example

```{r setup}
library(paulmisc)
library(ggplot2)
library(plotly)

# Generate example outbreak data
cases <- simulate_outbreak(n = 50, seed = 123)
```

First, let's create a basic interactive epidemic curve:

```{r basic-interactive, message=FALSE}
# Create ggplot
p <- ggplot(cases, aes(x = onset_date)) +
  geom_epicurve(fill = "steelblue") +
  labs(
    title = "Interactive Epidemic Curve",
    x = "Date of Onset",
    y = "Number of Cases"
  ) +
  theme_minimal()

# Convert to interactive plotly plot
ggplotly(p)
```

Hover over any square to see details!

## Interactive by Age Group

Now let's colour by age group and add custom tooltips:

```{r by-age-group, message=FALSE}
# Add custom tooltip text
cases$tooltip <- paste0(
  "<b>Case ", cases$case_id, "</b><br>",
  "Date: ", format(cases$onset_date, "%d %B %Y"), "<br>",
  "Age Group: ", cases$age_group, "<br>",
  "Sex: ", cases$sex, "<br>",
  "Setting: ", cases$setting, "<br>",
  "Outcome: ", cases$outcome
)

# Create plot with custom tooltips
p <- ggplot(cases, aes(x = onset_date, fill = age_group, text = tooltip)) +
  geom_epicurve() +
  scale_fill_brewer(palette = "Set2", name = "Age Group") +
  labs(
    title = "Interactive Epidemic Curve by Age Group",
    x = "Date of Onset",
    y = "Number of Cases"
  ) +
  theme_minimal()

# Convert to interactive, showing only our custom tooltip
ggplotly(p, tooltip = "text")
```

## Faceted Interactive Plots

Interactive plots work with faceting too:

```{r faceted, message=FALSE, fig.height=6}
p <- ggplot(cases, aes(x = onset_date, fill = sex, text = tooltip)) +
  geom_epicurve() +
  scale_fill_manual(
    values = c("Male" = "#0072B2", "Female" = "#D55E00"),
    name = "Sex"
  ) +
  facet_wrap(~ setting, ncol = 1, scales = "free_y") +
  labs(
    title = "Interactive Epidemic Curves by Setting",
    x = "Date of Onset",
    y = "Number of Cases"
  ) +
  theme_minimal()

ggplotly(p, tooltip = "text")
```

## Customising Tooltip Content

You have complete control over tooltip content. Use HTML formatting for rich tooltips:

```{r custom-tooltips, message=FALSE}
# Create highly customised tooltips with HTML formatting
cases$tooltip <- with(cases, paste0(
  "<b style='font-size:14px'>Case ", case_id, "</b><br>",
  "<hr style='margin:2px'>",
  "<i>Date:</i> ", format(onset_date, "%d %B %Y"), "<br>",
  "<i>Demographics:</i> ", age_group, ", ", sex, "<br>",
  "<i>Setting:</i> ", setting, "<br>",
  "<i>Outcome:</i> ", outcome
))

p <- ggplot(cases, aes(x = onset_date, fill = outcome, text = tooltip)) +
  geom_epicurve() +
  scale_fill_brewer(palette = "Pastel1", name = "Outcome") +
  labs(
    title = "Interactive Epidemic Curve by Outcome",
    x = "Date of Onset",
    y = "Number of Cases"
  ) +
  theme_minimal()

ggplotly(p, tooltip = "text")
```

## Tips for Interactive Visualisation

- **Simple conversion**: Just wrap your ggplot with `ggplotly()` for basic interactivity
- **Custom tooltips**: Add a `text` aesthetic and use `ggplotly(p, tooltip = "text")` to show only your custom tooltip
- **Tooltip formatting**: Use HTML tags (`<br>`, `<b>`, `<i>`, `<hr>`) for rich formatting
- **Performance**: Interactive plots with many cases (>500) may be slow. Consider filtering for large datasets
- **Mobile devices**: Interactive features work on touch devices - tap to see tooltips
- **Export**: Use plotly's built-in export button to save static images

## More Information

For more details on plotly customisation, see the [plotly for R documentation](https://plotly-r.com/).
