---
title: "Interactive Epidemic Curves with Plotly"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Interactive Epidemic Curves with Plotly}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Overview

The `geom_epicurve()` function can be combined with [plotly](https://plotly.com/r/) to create interactive epidemic curves with custom tooltips. This allows users to hover over individual case squares to see detailed information.

## Basic Interactive Example

```{r setup}
library(paulmisc)
library(ggplot2)
library(plotly)

# Generate example outbreak data
cases <- simulate_outbreak(n = 50, seed = 123)
```

First, let's create a basic interactive epidemic curve:

```{r basic-interactive, message=FALSE}
# Add custom tooltip text to the data
cases$tooltip <- paste0(
  "Case ID: ", cases$case_id, "\n",
  "Date: ", cases$onset_date, "\n",
  "Age: ", cases$age_group, "\n",
  "Sex: ", cases$sex
)

# Create plot with text aesthetic for tooltips
p <- ggplot(cases, aes(x = onset_date, text = tooltip)) +
  geom_epicurve(fill = "steelblue") +
  labs(
    title = "Interactive Epidemic Curve",
    x = "Date of Onset",
    y = "Number of Cases"
  ) +
  theme_minimal()

# Convert to interactive plotly plot
ggplotly(p, tooltip = "text")
```

Hover over any square to see the details for that individual case!

## Interactive by Age Group

Now let's colour by age group and customise the tooltips further:

```{r by-age-group, message=FALSE}
# Enhanced tooltip with more detail
cases$tooltip <- paste0(
  "<b>Case ", cases$case_id, "</b>\n",
  "Date: ", format(cases$onset_date, "%d %B %Y"), "\n",
  "Age Group: ", cases$age_group, "\n",
  "Sex: ", cases$sex, "\n",
  "Setting: ", cases$setting, "\n",
  "Outcome: ", cases$outcome
)

# Create plot with fill by age group
p <- ggplot(cases, aes(x = onset_date, fill = age_group, text = tooltip)) +
  geom_epicurve() +
  scale_fill_brewer(palette = "Set2", name = "Age Group") +
  labs(
    title = "Interactive Epidemic Curve by Age Group",
    x = "Date of Onset",
    y = "Number of Cases"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")

# Convert to interactive
ggplotly(p, tooltip = "text")
```

## Interactive Faceted Plot

Interactive plots work with faceting too:

```{r faceted, message=FALSE, fig.height=6}
# Create faceted plot
p <- ggplot(cases, aes(x = onset_date, fill = sex, text = tooltip)) +
  geom_epicurve() +
  scale_fill_manual(
    values = c("Male" = "#0072B2", "Female" = "#D55E00"),
    name = "Sex"
  ) +
  facet_wrap(~ setting, ncol = 1, scales = "free_y") +
  labs(
    title = "Interactive Epidemic Curves by Setting",
    x = "Date of Onset",
    y = "Number of Cases"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")

# Convert to interactive
ggplotly(p, tooltip = "text")
```

## Customising Tooltip Content

The `text` aesthetic gives you complete control over tooltip content. You can include any information from your data and format it however you like:

```{r custom-tooltips, message=FALSE}
# Create highly customised tooltips with HTML formatting
cases$tooltip <- with(cases, paste0(
  "<b style='font-size:14px'>Case ", case_id, "</b><br>",
  "<hr style='margin:2px'>",
  "<i>Date:</i> ", format(onset_date, "%d %B %Y"), "<br>",
  "<i>Demographics:</i> ", age_group, ", ", sex, "<br>",
  "<i>Setting:</i> ", setting, "<br>",
  "<i>Outcome:</i> ", outcome
))

p <- ggplot(cases, aes(x = onset_date, fill = outcome, text = tooltip)) +
  geom_epicurve() +
  scale_fill_brewer(palette = "Pastel1", name = "Outcome") +
  labs(
    title = "Interactive Epidemic Curve by Outcome",
    x = "Date of Onset",
    y = "Number of Cases"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")

ggplotly(p, tooltip = "text")
```

## Tips for Interactive Visualisation

- **Performance**: Interactive plots with many cases (>500) may be slow. Consider aggregating or filtering for large datasets.
- **Tooltip formatting**: Use `\n` for line breaks in plain text, or HTML tags (`<br>`, `<b>`, `<i>`) for richer formatting.
- **Mobile devices**: Interactive features work on touch devices - tap to see tooltips.
- **Export**: Use plotly's built-in export button to save static images of your interactive plots.

## More Information

For more details on plotly customisation, see the [plotly for R documentation](https://plotly-r.com/).
